### 2021-05-11 proposal/proposal.tex
updated proposal

Graph and network data have become increasingly more widespread in
various fields including sociology, neuroscience, biostatistics, and
computer science. This has resulted in certain challenges for
researchers, as most traditional statistical and machine learning
methods are incompatible with graphs and instead require the data to
exist in a subset of Euclidean space. This includes clustering and
community detection, which are the typical problems of interest when
working with graphs. Common clustering methods often involve calculating
some central or representative point around which the data belonging to
that cluster lie (e.g., Lloyd's algorithm for \(K\)-means clustering
\cite{1056489}, Gaussian Mixture Models
\cite{doi:10.1198/016214502760047131}). Because these methods involve
computing summary statistics within each cluster, they cannot be applied
directly to graphs as community detection algorithms.

Some proposed methods for unifying graph community detection with
traditional clustering techniques involve embedding the graph into
Euclidean space, of which the most popular is Spectral Clustering
\cite{DBLP:journals/corr/abs-0711-0189}. The Random Dot Product Graph
(RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot
Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
this further by explicitly constructing generative models such that
latent positions in Euclidean space are used to induce graphs. A
community detection method motivated by this would involve learning the
latent positions given an observed graph and then learning the community
labels given the latent positions.

@@

Let \(G = (V, E)\) be an undirected, unweighted graph with no self-loops
with \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
edge between vertices \(i\) and \(j\) and \(A_{ij} = 0\) otherwise.
Because \(G\) is symmetric and contains no self-loops,
\(A_{ij} = A_{ji}\) and \(A_{ii} = 0\) for \(i, j \in [n]\). We further
restrict our analyses to independent Bernoulli graphs. Let
\(P \in [0, 1]^{n \times n}\) be a symmetric matrix of edge
probabilities. Graph \(G\) is sampled by
\(A_{ij} \stackrel{\text{indep}}{\sim}P_{ij}\) for each
\(0 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)). Denote
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
as the sample \(x_1, ..., x_n \in \mathbb{R}^d\), and denote
\(z_1, ..., z_n \in [K]\) as their corresponding (hidden) labels.

@@

Generative models for symmetric Bernoulli graphs involve defining the
edge probability matrix \(P\) whose \(ij\)\textsuperscript{th} entry is
the probability of an edge between vertices \(i\) and \(j\) for each
\(i, j \in [n]\). Community detection and recovery starts by restricting
the generative model such that for each pair of vertices, the
probability of an edge between the vertices is conditioned on the labels
of the vertices. One such model is the Stochastic Block Model (SBM)
\cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
each vertex belonging to one community, the SBM restricts \(P\) to
\(K (K + 1) / 2\) unique entries such that \(P_{ij} = B_{z_i z_j}\)
where \(B \in [0, 1]^{K \times K}\) with entries \(B_{kl}\) denoting the
edge probability of each vertex in community \(k\) having an edge with
each vertex in community \(l\). The homogeneous SBM further restricts
\(P\) to two unique entries such that \(P_{ij} = p\) if \(z_i = z_j\)
and \(P_{ij} = q\) otherwise. Multiple generalizations of the SBM have
been introduced since, including the Degree Corrected Block Model (DCBM)
and the Popularity Adjusted Block Model (PABM). Like the SBM, these
models involve edge probability matrix \(P\) that is restricted to rank
\(d < n\) in part based on community labels.

@@

The RDPG starts with points in latent space
\(X \in \mathcal{X} \subset \mathbb{R}^d\) such that
\(\forall x, y \in \mathcal{X}\), \(x^\top y \in [0, 1]\). \(P\) is then
constructed as \(P = X X^\top\), and graph \(G\) with adjacency matrix
\(A\) is drawn from \(P\). We provide a more formal definition of the
RDPG and GRDPG below.

@@

\citeauthor{athreya2017statistical} showed that under mild conditions,
if \((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),

@@

It is straightforward to show that all Bernoulli graphs with positive
semidefinite \(P\) are special cases of the RDPG, which is a special
case of the GRDPG, and all graphs generated by \(P\) are special cases
of the GRDPG. This includes the SBM, DCBM, and PABM.

@@

\hypertarget{subspace-clustering}{%
\subsection{Subspace Clustering}\label{subspace-clustering}}

Subspace clustering, which assumes that points in \(\mathbb{R}^d\) each
lie on one of \(K\) subspaces of \(\mathbb{R}^d\), is an approach that
has found a wide range of uses by the Statistics and Machine Learning
communities, particularly within the field of Computer Vision
\cite{5206547}.

Of particular interest is Sparse Subspace Clustering (SSC), which is
performed by solving an optimization problem for each observed point in
a sample. Given \(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
collected into matrix
\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
perfectly on one of \(K\) subspaces, \(B\) is sparse such that
\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
and if the number of subgraphs is exactly \(K\), each subgraph maps onto
a subspace.

@@

\hypertarget{subspace-clustering-1}{%
\subsubsection{Subspace Clustering}\label{subspace-clustering-1}}

As discussed in \S 2.2, if the latent positions of a RDPG or GRDPG model
are such that they lie on a small number of subspaces, ASE followed by
SSC may be able to identify whether vertices of the graph \(v_i, v_j\)
belong to the same subspace or to different subspaces, and one such
model that is consistent with this construction is the PABM.

@@

This leads to the following theorem.

@@

If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), Theorem \ref{osc}
and results from \citeauthor{rubindelanchy2017statistical} together
imply \(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} 0\), leading to the
following result:

@@

@@

\hypertarget{manifold-clustering}{%
\subsection{Manifold Clustering}\label{manifold-clustering}}

We would like to extend subspace clustering to manifold clustering. The
problem setup is as follows: Suppose that in the latent space
\(\mathcal{X} \subset \mathbb{R}^d\), sample \(X\) of \(n\) points lie
on a union of \(K\) disjoint manifolds with each manifold corresponding
to a community. If \(A \sim \text{RDPG}(X)\), we wish to recover the
community labels (up to permutation) from \(A\).

Similarly, suppose that probability distribution \(F\) is described as
follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Define functions \(f_1, ..., f_K\) such that
  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_k(t) \neq f_l(t)\)
  \(\forall k, l \in [K]\) and \(t \in [0, 1]\).
\item
  Sample labels
  \(z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)\).
\item
  Sample \(t_1, ..., t_n \stackrel{iid}{\sim} D\) where \(D\) has
  support \([0, 1]\).
\item
  Set latent positions \(x_i = f_{z_i}(t_i)\) and
  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\end{enumerate}

@@

\hypertarget{affine-subspaces}{%
\subsubsection{Affine Subspaces}\label{affine-subspaces}}

@@

If instead of sampling uniformly from line segments of unit length, we
sample uniformly from a 1 dimensional manifolds of unit length, the
above property still holds. Let
\(U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)\) and
\(f : [0, 1] \mapsto \mathbb{R}^d\) be a smooth function such that
\(\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|\). Then
\(U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
\(P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
If the shortest distance between the two manifolds defined by \(f_1\)
and \(f_2\) with the same restriction is \(a\), then the same \(N\) as
before is sufficient, although perhaps a more lenient lower bound can be
derived based on the shape of \(f_i(\cdot)\).

>>>>

Graph and network data have become increasingly widespread in various
fields including sociology, neuroscience, biostatistics, and computer
science. This has resulted in various challenges for researchers who
rely on traditional statistical and machine learning methods, many of
which are incompatible with graph data and instead require the data to
exist as feature vectors in Euclidean space. Such challenges often
involve clustering and community detection. Common clustering methods
typically involve calculating some central or representative point for
each cluster around which the data belonging to that cluster lie (e.g.,
Lloyd's algorithm for \(K\)-means clustering \cite{1056489}, Gaussian
Mixture Models \cite{doi:10.1198/016214502760047131}). Because these
methods involve computing summary statistics within each cluster, such
as the sample average, they cannot be applied directly to graphs,
necessitating methods for transforming the graph data into feature
vectors.

One family of methods for unifying graph community detection with
traditional clustering techniques is Spectral Clustering
\cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the
graph into Euclidean space, followed by applying a popular clustering
algorithm such as \(K\)-means clustering. The Random Dot Product Graph
(RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot
Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take
this further by explicitly constructing generative models such that
latent positions in Euclidean space are used to induce graphs. A
community detection algorithm motivated by this may involve learning the
latent positions given an observed graph and then learning the community
labels given the latent positions.

@@

Let \(G = (V, E)\) be an undirected, unweighted graph with no self-loops
with \(n\) vertices. Denote \(A \in \{0, 1\}^{n \times n}\) as the
adjacency matrix for \(G\) such that \(A_{ij} = 1\) if there exists an
edge between vertices \(i\) and \(j\) and \(A_{ij} = 0\) otherwise.
Because \(G\) is symmetric and contains no self-loops,
\(A_{ij} = A_{ji}\) and \(A_{ii} = 0\) for \(i, j \in [n]\). We further
restrict our analyses to independent Bernoulli graphs. Let
\(P \in [0, 1]^{n \times n}\) be a symmetric matrix of edge
probabilities. Graph \(G\) is sampled from \(P\) by drawing
\(A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})\) for
each \(0 \leq i < j \leq n\) (\(A_{ji} = A_{ij}\) and \(A_{ii} = 0\)).
Finally, we denote
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}\)
as the sample \(x_1, ..., x_n \in \mathbb{R}^d\), and denote
\(z_1, ..., z_n \in [K]\) as their corresponding (hidden) labels.

@@

Generative models for symmetric Bernoulli graphs involve defining the
edge probability matrix \(P\) whose \(ij\)\textsuperscript{th} entry is
the probability of an edge between vertices \(i\) and \(j\) for each
\(i, j \in [n]\). In order to motivate community detection methods, we
restrict the generative model such that for each pair of vertices, the
probability of an edge between the vertices is conditioned on the labels
of the vertices. One such model is the Stochastic Block Model (SBM)
\cite{doi:10.1080/0022250X.1971.9989788}: Given \(K\) communities and
each vertex belonging to one community, the SBM restricts \(P\) to
\(K (K + 1) / 2\) unique entries such that \(P_{ij} = B_{z_i z_j}\) and
\(B \in [0, 1]^{K \times K}\) with entries \(B_{kl}\) denoting the edge
probability of each vertex in community \(k\) having an edge with each
vertex in community \(l\). The homogeneous SBM further restricts \(P\)
to two unique entries such that \(P_{ij} = p\) if \(z_i = z_j\) and
\(P_{ij} = q\) otherwise. Multiple generalizations of the SBM have been
introduced since, including the Degree Corrected Block Model (DCBM) and
the Popularity Adjusted Block Model (PABM). Like the SBM, these models
involve edge probability matrix \(P\) that is restricted to rank
\(d < n\) in part based on community labels.

@@

Like the SBM, DCBM, and PABM, the RDPG and GRDPG are generative models
for graphs involving an edge probability matrix \(P\). The RDPG starts
with points in latent space \(X \in \mathcal{X} \subset \mathbb{R}^d\)
such that \(\forall x, y \in \mathcal{X}\), \(x^\top y \in [0, 1]\).
\(P\) is then constructed as \(P = X X^\top\), and graph \(G\) with
adjacency matrix \(A\) is drawn from \(P\). We provide a more formal
definition of the RDPG and GRDPG below.

@@

\citet{athreya2017statistical} showed that under mild conditions, if
\((A_n, X_n) \sim \text{RDPG}(F, n)\) and \(\hat{X}_n\) is the scaled
ASE of \(A_n\), for some sequence of orthogonal matrices \(W_n\),

@@

It is straightforward to show that all Bernoulli graphs with positive
semidefinite \(P\) are special cases of the RDPG, which is a special
case of the GRDPG, and all graphs generated by \(P\) are special cases
of the GRDPG. This includes the SBM, DCBM, and PABM. In the following
example, we show that the ASE of the SBM has a very particular form.

@@

@@

As discussed in \S 2.1, all Bernoulli graph models can be expressed as a
RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
with a very specific structure in the latent space. A similar result has
been shown for the DCBM. We will now extend this to the PABM.

@@

In Theorem \ref{theorem2}, we showed that a possible latent
configuration for a \(K\)-community PABM under the GRDPG framework
consists of latent positions in \(\mathbb{R}^{K^2}\) such that each
community corresponds to a \(K\)-dimensional subspace such that each
subspace is orthogonal to the others.

@@

If \(\hat{V}\) is the \emph{unscaled} ASE of \(A\), then results from
\citeauthor{rubindelanchy2017statistical} imply
\(n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top\). Then for
vertices \(i\) and \(j\) belonging to separate communities, the
\(ij\)\textsuperscript{th} entry of \(n \hat{V} \hat{V}^\top\)
approaches 0 with probability 1:

@@

\hypertarget{subspace-clustering}{%
\subsubsection{Subspace Clustering}\label{subspace-clustering}}

In addition to OSC, since the communities are subspaces in the latent
space, we can leverage existing methods for subspace clustering. Of
particular interest is Sparse Subspace Clustering (SSC), which is
performed by solving an optimization problem for each observed point in
a sample. Given \(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
collected into matrix
\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
perfectly on one of \(K\) subspaces, \(B\) is sparse such that
\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
and if the number of subgraphs is exactly \(K\), each subgraph maps onto
a subspace.

@@

\hypertarget{proposed-generative-graph-models}{%
\subsection{Proposed Generative Graph
Models}\label{proposed-generative-graph-models}}

In the previous sections, we connected well-known and highly structured
generative graph models to the RDPG and GRDPG to show that highly
structured latent configurations generate graphs consistent with these
models: The latent space for the SBM consists of \(K\) point masses, the
latent space for the DCBM consists of \(K\) rays emanating from the
origin, and the latent space for the PABM consists of \(K\)
\(K\)-dimensional subspaces. In the following sections, we explore
additional structured latent configurations corresponding to community
structure and develop methods for community detection based on the
consistency of the ASE and the structural forms of the latent
configurations.

The general structure of interest can be described as follows: Suppose
that in the latent space \(\mathcal{X} \subset \mathbb{R}^d\), sample
\(X\) of \(n\) points lie on a union of \(K\) disjoint manifolds with
each manifold corresponding to a community. If
\(A \sim \text{RDPG}(X)\), we wish to recover the community labels (up
to permutation) from \(A\).

Equivalently, suppose that probability distribution \(F\) is described
as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Define functions \(f_1, ..., f_K\) such that
  \(f_k : [0, 1] \mapsto \mathcal{X}\) and \(f_k(t) \neq f_l(t)\)
  \(\forall k, l \in [K]\) and \(t \in [0, 1]\).
\item
  Sample labels
  \(z_1, ..., z_n \stackrel{\text{iid}}{\sim}Categorical(\pi_1, ..., \pi_K)\).
\item
  Sample \(t_1, ..., t_n \stackrel{\text{iid}}{\sim}D\) where \(D\) has
  support \([0, 1]\).
\item
  Set latent positions \(x_i = f_{z_i}(t_i)\) and
  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\end{enumerate}

@@

\hypertarget{affine-segments}{%
\subsubsection{Affine Segments}\label{affine-segments}}

@@

If instead of sampling uniformly from line segments of unit length, we
sample uniformly from a 1 dimensional manifolds of unit length, the
above property still holds. Let
\(U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)\) and
\(f : [0, 1] \mapsto \mathbb{R}^d\) be a smooth function such that
\(\int_u^v \sqrt{\sum_k (df_k / dt)^2} dt = \|u - v\|\). Then
\(U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|\), so
\(P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)\).
If the shortest distance between the two manifolds defined by \(f_1\)
and \(f_2\) with the same restriction is \(a\), then the same \(N\) as
before is sufficient, although perhaps a more lenient lower bound can be
derived based on the shape of \(f_k(\cdot)\).
